---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This document is the submission for Peer Assignment 1 of the Coursera Data Science course *Reproducible Research*. It was forked from the GitHub repository [RepData_PeerAssessment1](https://github.com/rdpeng/RepData_PeerAssessment1) and extended as specified in the [exercise description](https://class.coursera.org/repdata-011/human_grading/view/courses/973512/assessments/3/submissions). 

## Loading and preprocessing the data

Load the data in a CSV file wrapped up in a zip file. As the date column will be a factor convert it to a POSIX date field. The first few rows of the data frame are displayed for sanity checking.

```{r}
# read the CSV file within the ZIP file
activity <- read.csv (unz ("activity.zip", "activity.csv"))

# convert the date to a POSIX date column
activity$date <- as.POSIXct (activity$date)

# examine the first few rows
head (activity)
```

## What is mean total number of steps taken per day?

I am using Hadley Wickham's library *dplyr* to massage the activity data into the required form. The NA measurements are filtered out and the steps are summed up for each day. A histogram of the steps per day (bin width of 5000 steps) shows the distribution.

```{r}
library(ggplot2)
suppressPackageStartupMessages (library (dplyr, quietly=TRUE))

# use dplyr's pipe operator to filter and summarize
stepsByDate <-
  activity                      %>% # from activity ...
  filter (!is.na (steps))       %>% # take only valid step counts ...
  group_by (date)               %>% # group it by the date ...
  summarize (nSteps=sum (steps))    # and total up the steps

# plot a histogram of the steps
qplot (stepsByDate$nSteps,
       xlab="Distribution of steps per day", ylab="Count",
       binwidth=5000)
```

The mean and median values of the steps per day are almost identical.

```{r}
# write out the mean of the steps
mean (stepsByDate$nSteps)

# write the median of the steps
median (stepsByDate$nSteps)
```

## What is the average daily activity pattern?

The time series plot of the average step count by interval showing the activity over a 24-hour period. One can discern the diurnal cycle: sleep, followed by high activity in the morning, less intense movement in the afternoon and a quietening down towards the evening.

```{r}
meanStepsByInterval <-
  activity                      %>% # from activity ...
  filter (!is.na (steps))       %>% # take only the valid step counts ...
  group_by (interval)           %>% # group it by the interval ...
  summarize (nSteps=mean (steps))   # and calculate the average #steps

qplot (meanStepsByInterval$interval,
       meanStepsByInterval$nSteps,
       xlab="Interval",
       ylab="Average number of steps",
       geom="line") +
   geom_hline (aes (yintercept=mean (meanStepsByInterval$nSteps)),
               color="yellow",
               alpha=0.8)
```

The most active interval (ie the 5-minute period with the most number of steps) is shown together the highest step count.


```{r}
highestStepCount <- max (meanStepsByInterval$nSteps)
meanStepsByInterval [meanStepsByInterval$nSteps == highestStepCount, ]
```

## Imputing missing values

The total number of missing values for the steps field:

```{r}
sum (is.na (activity$steps))
```

The strategy chosen for imputing missing values is to replace NA's with the average number of steps for that time interval.

```{r}
# make a new table from the original activity with a new field called nSteps that
# contains the average of that interval
cleanedActivity <- merge (meanStepsByInterval, activity, by="interval")
cleanedActivity$steps <-
  ifelse (is.na (cleanedActivity$steps),
          round (cleanedActivity$nSteps),
          cleanedActivity$steps)

# as done earlier, total the steps of each day
stepsByDate <-
  cleanedActivity               %>% # from activity with imputed missing data ...
  group_by (date)               %>% # group it by the date ...
  summarize (nSteps=sum (steps))    # and total up the steps

# plot a histogram of the steps
qplot(stepsByDate$nSteps,
      xlab="Distribution of steps per day with imputed missing values",
      ylab="Count", binwidth=5000)

# write out the new mean of the steps after cleaning the data
mean (stepsByDate$nSteps)

# the median of the steps after cleaning the data
median (stepsByDate$nSteps)

```

As one can see, there is a small change in the median after imputing missing values; the mean is almost the same as before.

From the histogram one can see that total daily number of steps has increased when one compares it with the previous one.


## Are there differences in activity patterns between weekdays and weekends?

The date is used to ascertain if it was a weekeday or a weekend and based on this factor two plots are made with the enhanced data.

```{r}
# this function returns either "weekday" or "weekend" depending on the date given to it.
weekdayType <- function (x) {
  listOfWeekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
  ifelse (weekdays (x) %in% listOfWeekdays, "weekday", "weekend")
}

# add a new column for the weekday type
cleanedActivity$wd <- as.factor (weekdayType (cleanedActivity$date))

# calculate the mean number of steps in each interval for weekdays and weekends
meanStepsByInterval <-
  cleanedActivity               %>% # from activity with imputed missing data ...
  group_by (interval, wd)       %>% # group it by the date ...
  summarize (nSteps=mean (steps))   # and get the average #steps

# plot a faceted graph of the mean number of steps
qplot (interval, nSteps, data=meanStepsByInterval,
       xlab="Interval", ylab="Average number of steps",
       geom="line",
       facets=wd ~ .) 
```


The two plots show that the weekday and weekend activities are noticeably  different.
